Clone and graft:\\ Testing scientific applications as they are built

### 2015-01-22 turcksin-spring-mechanics/paper.tex
fix typos and small edits

\author{Bruno Turcksin\footnote{Department of Mathematics, Texas A\&M
    University, College Station, TX 77843-3368, USA, \url{???}}
\and
   Timo Heister\footnote{Mathematical Sciences, O-110 Martin Hall, Clemson University,
     Clemson, SC 29634-095, USA, \url{heister@clemson.edu}}
\and
   Wolfgang Bangerth\footnote{Corresponding author. Department of Mathematics, Texas A\&M
     University, College Station, TX 77843-3368, USA, \url{bangerth@math.tamu.edu}}}

@@

\section{\aspect{}: the Advanced Solver for Problems in Earth ConvecTion}
\label{sec:aspect}

\begin{wrapfigure}{R}{0.44\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.42\textwidth,height=0.42\textwidth]{figures/aspect.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Snapshot of the temperature field of an \aspect{} simulation
    showing convection in the Earth mantle.}
  \vspace*{-3mm}
  \label{fig:aspect}
\end{wrapfigure}
The code that made us think about how to test whole applications is called
\aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
\cite{KHB12,aspectmanual,aspectweb}. It is a program that simulates how material moves around the Earth
mantle (i.e., the region between the metallic core of Earth and the plates at
the surface on which we live). While the material in the mantle is solid rock,
it is hot and under enormous pressure and can deform with velocities of a few
centimeters per year. On time scales of millions of years, it therefore
behaves like a fluid. Since it is heated from below and cooled from above, it
shows the same kind of behavior as a pot on the stove: blobs of hot material
rise up, cool at the surface, and blobs of cool material fall down. In other
words, it convects. The details of trying to simulate this on a computer go
beyond what we want to discuss here, but it is worth showing a picture and
linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
\url{http://www.youtube.com/embed/EJJ6f4hmDPU}, of only because they are pretty.

@@

Let's implement a program that can solve this problem. Because we want it
to be simple and accessible, we're going to use Python as it's widely used and
because it comes with a number of tools that are going to make our life
simpler by keeping the program short.%
\footnote{In order to run this program, you will have to have the Python
  NumPy, SciPy, and MatPlotLib packages installed on your system, along with
  the basic Python interpreter.}
The program we'll show you in a second reads its input parameters from a file in JSON format%
\footnote{JSON is not the prettiest format to write input files in and we
  don't recommend you use it for your programs (it is typically intended for
  exchange of data between programs, not between humans and computers). But it
  will serve for our purposes here primarily because Python has a built-in
  parser for it that makes our program pleasantly free of the tedium of
  parsing input files.} (see
\url{http://en.wikipedia.org/wiki/JSON}).
Here is an example that you can find in the file \texttt{tests/testcase-1.json}:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-1.json}
In JSON, the name of a parameter is to the left of the colon and its value is
to the right; name/value pairs are separated by commas. The value of a
parameter can be a plain number, or it can be an array if enclosed in
brackets. Arrays can also be nested, which is what we use here for the initial
positions and velocities of our two bodies, each of which lives in
three-dimensional space. With this explanation, it is clear that the input file
specifies properties of the two masses as $\mathbf x_{1,0}=(1,2,3)^T, \mathbf
x_{2,0}=(4,5,6)^T, \mathbf v_{1,0}=(0,0,0)^T, \mathbf v_{2,0}=(1,1,1)^T,
m_1=13.5, m_2=29.75$ and $D=42, L=2.25$ for the spring that connects them, 
thereby completely determining everything in equations
\eqref{eq:ode}--\eqref{eq:ic2}.

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:

@@

Implementing all of this leads to the following extension of the program shown
above%
\footnote{This is the file \texttt{spring.py} in the directory corresponding to
this paper}, with the new parts in lines 19--20 and 35--36:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}

@@

From experience, it turns out that this is not often a real problem. The vast
majority of patches concern just small, isolated pieces of the software and in
those cases, only one or a few testcases will start failing. It is then easy
to identify the simplest one of these (assuming that they were given useful
names, see above in Section~\ref{sec:many-tests}), and that is the one we
will then use for debugging. If one breaks fundamental functionality somewhere
in the core of the software, a large number of tests will start to fail
but in those cases you will either be able to quickly find out what is
happening, or be able to debug the problem by choosing \textit{any} simple
test that was among those that broke. In summary, from a practical
perspective, we have found that it is not usually a problem to have testcases
that check overlapping parts of the software. Rather, we have come to the
conclusion that having as many tests as possible, regardless of overlap.

@@

\paragraph*{Acknowledgments}
B.~Turcksin and W.~Bangerth were partially supported by the National Science
Foundation under award OCI-1148116 as part of the Software Infrastructure for
Sustained Innovation (SI2) program; and by the Computational
Infrastructure in Geodynamics initiative (CIG), through the National Science
Foundation under Award No.~EAR-0949446 and The University of California --
Davis.
\marginpar{Timo -- please update.}

>>>>

\author{Bruno Turcksin\footnote{Department of Mathematics, Texas A\&M
    University, College Station, TX 77843-3368, USA, \url{???}}
\and
   Timo Heister\footnote{Mathematical Sciences, O-110 Martin Hall, Clemson University,
     Clemson, SC 29634-0975, USA, \url{heister@clemson.edu}}
\and
   Wolfgang Bangerth\footnote{Corresponding author. Department of Mathematics, Texas A\&M
     University, College Station, TX 77843-3368, USA, \url{bangerth@math.tamu.edu}}}

@@

\section{\aspect{}: the Advanced Solver for Problems in Earth's ConvecTion}
\label{sec:aspect}

\begin{wrapfigure}{R}{0.44\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.42\textwidth,height=0.42\textwidth]{figures/aspect.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Snapshot of the temperature field of an \aspect{} simulation
    showing convection in the Earth mantle.}
  \vspace*{-3mm}
  \label{fig:aspect}
\end{wrapfigure}
The code that made us think about how to test whole applications is called
\aspect{} -- short for the Advanced Solver for Problems in Earth's ConvecTion
\cite{KHB12,aspectmanual,aspectweb}. It is a program that simulates how material moves around the Earth
mantle (i.e., the region between the metallic core of Earth and the plates at
the surface on which we live). While the material in the mantle is solid rock,
it is hot and under enormous pressure and can deform with velocities of a few
centimeters per year. On time scales of millions of years, it therefore
behaves like a fluid. Since it is heated from below and cooled from above, it
shows the same kind of behavior as a pot on the stove: blobs of hot material
rise up, cool at the surface, and blobs of cool material fall down. In other
words, it convects. The details of trying to simulate this on a computer go
beyond what we want to discuss here, but it is worth showing a picture and
linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
\url{http://www.youtube.com/embed/EJJ6f4hmDPU}, if only because they are pretty.

@@

Let's implement a program that can solve this problem. Because we want it
to be simple and accessible, we're going to use Python as it's widely used and
because it comes with a number of tools that are going to make our life
simpler by keeping the program short.%
\footnote{In order to run this program, you will have to have the Python
  NumPy, SciPy, and matplotlib packages installed on your system, along with
  the basic Python interpreter.}
The program we'll show you in a second reads its input parameters from a file in JSON format%
\footnote{While certainly more sane compared to XML, JSON is not the prettiest format to write input files in and we
  don't recommend you use it for your programs (it is typically intended for
  exchange of data between programs, not between humans and computers). But it
  will serve for our purposes here primarily because Python has a built-in
  parser for it that makes our program pleasantly free of the tedium of
  parsing input files.} (see
\url{http://en.wikipedia.org/wiki/JSON}).
Here is an example that you can find in the file \texttt{tests/testcase-1.json}:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-1.json}
In JSON, the name of a parameter is to the left of the colon and its value is
to the right; name/value pairs are separated by commas. The value of a
parameter can be a plain number, or it can be an array if enclosed in
brackets. Arrays can also be nested, which is what we use here for the initial
positions and velocities of our two bodies, each of which lives in
three-dimensional space. With this explanation, it is clear that the input file
specifies properties of the two masses as $\mathbf x_{1,0}=(1,2,3)^T, \mathbf
x_{2,0}=(4,5,6)^T, \mathbf v_{1,0}=(0,0,0)^T, \mathbf v_{2,0}=(1,1,1)^T,
m_1=13.5, m_2=29.75$ and $D=42, L=2.25$ for the spring that connects them, 
thereby completely determining everything in equations
\eqref{eq:ode}--\eqref{eq:ic2}.

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables for the velocities $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:

@@

Implementing all of this leads to the following extension of the program shown
above%
\footnote{This is the file \texttt{spring.py} in the directory corresponding to
this paper.}, with the new parts in lines 19--20 and 35--36:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}

@@

From experience, it turns out that this is not often a real problem. The vast
majority of patches concern just small, isolated pieces of the software and in
those cases, only one or a few testcases will start failing. It is then easy
to identify the simplest one of these (assuming that they were given useful
names, see above in Section~\ref{sec:many-tests}), and that is the one we
will then use for debugging. If one breaks fundamental functionality somewhere
in the core of the software, a large number of tests will start to fail
but in those cases you will either be able to quickly find out what is
happening, or be able to debug the problem by choosing \textit{any} simple
test that was among those that broke. In summary, from a practical
perspective, we have found that it is not usually a problem to have testcases
that check overlapping parts of the software. Rather, we have come to the
conclusion that it is best to have as many tests as possible, regardless of overlap.

@@

\paragraph*{Acknowledgments}
B.~Turcksin and W.~Bangerth were partially supported by the National Science
Foundation under award OCI-1148116 as part of the Software Infrastructure for
Sustained Innovation (SI2) program; and by the Computational
Infrastructure in Geodynamics initiative (CIG), through the National Science
Foundation under Award No.~EAR-0949446 and The University of California --
Davis.

T.~Heister was partially supported by the Computational
Infrastructure in Geodynamics initiative (CIG), through the National Science
Foundation under Award No.~EAR-0949446 and The University of California --
Davis.
